Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
r overview

\lstset{language=R}
\definecolor{LightGray}{rgb}{.92,.92,.92}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}
\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  tabsize=2,
  belowskip=0.5em,
  backgroundcolor=\color{LightGray},
  basicstyle=\small\ttfamily,
  emphstyle=,
  keywordstyle=,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  numbers=left,
  showstringspaces=false
}
\lstdefinestyle{R}{ %
  language=R,
  morekeywords={assign, delayedAssign},
  deletekeywords={env, equal, c, runif, trace, args, exp, t, all},
  breaklines=true
}
\lstdefinestyle{Rin}{ %
  style=R,
  breaklines=false
}

@@

%
\subsection{The R Programming Language}

@@

In R, functions can be called with named parameters, arguments can have
default values, and R also support variable argument lists.  To illustrate
all of this in a single example, consider:

\begin{lstlisting}
f <- function(x, ..., y=3) x + y
\end{lstlisting}

\noindent
Function \k{f} can be called with a single argument \code{f(3)}, with named
argument \code{f(y=4,x=2)} and with a variable number of arguments,
\code{f(1,2,3,4,y=5)}, all of these calls will return \code{6}.

R has a number of features that are not crucial to the present
discussion. We will mention some of them here for completeness.  In R, data
structures are reference counted and have copy-on-write semantics, thus the
assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
the reference count on that object is 1.  This semantics gives R a
functional flavor while allowing updating in place within loops (the first
update copies, subsequent updates are performed on the copy). Arguments to
functions are evaluated only when needed, they are bundled in so-called
promises which package the original expression (as an abstract syntax tree,
or AST), its environment as well as the result of evaluating the
expression. Promises can be leveraged for meta-programming as it is possible
to retrieve the text of a promise and evaluate it in a different
environment.

\subsubsection{Types of Data}
\label{subsubsec:backgroundtypes}

Before attempting to define a type system for R, we should understand the
different kinds of values that programs operate on.  As we will see
different notions of type may emerge depending on how granular we want to
be.

\renewcommand{\k}[1]{{\tt #1}\xspace}

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{types} lists all of the builtin types that are
provided by the language. They are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R. But, in many
context a \k{logical} will convert to \k{integer}, and an \k{integer} will
convert to \k{double}.  Some off conversions can occur in corner cases, such
as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
vector, as the double is converted to an integer. R does not distinguish
between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}. Finally all
vectorized data types have a distinguished missing value denoted by
\code{NA}. The default type of \code{NA} is \k{logical}. We can see that
\code{typeof(NA)=="logical"}, but NA inhabits every type:
\code{typeof(c(1,NA)[2])=="double"}.

@@

Scalar data types include the distinguished \k{NULL} value, which is also of
type \k{NULL}, instance of classes written using the S4~\footnote{put footnote on S4 here} object system,
closures and environments.  The implementation of R has a number of other
types that are mostly not used by user code, they are listed in
Figure~\ref{types} for reference.

@@

%
%
\section{Project Corpus}\label{sec:corpus}

>>>>

\lstset{language=R}
\definecolor{LightGray}{rgb}{.92,.92,.92}
\definecolor{Gray}{rgb}{.3,.3,.3}
\definecolor{DarkGray}{rgb}{.5,.5,.5}
\lstset{ %
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
  tabsize=2,
  belowskip=0.5em,
  backgroundcolor=\color{LightGray},
  basicstyle=\small\ttfamily,
  emphstyle=,
  keywordstyle=,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
%  numbers=left,
  showstringspaces=false
}
\lstdefinestyle{R}{ %
  language=R,
  morekeywords={assign, delayedAssign},
  deletekeywords={env, equal, c, runif, trace, args, exp, t, all},
  breaklines=true
}
\lstdefinestyle{Rin}{ %
  style=R,
  breaklines=false
}
\renewcommand{\k}[1]{{\tt #1}\xspace}

@@

\subsection{The R Programming Language} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In this paper, we will attempt infer type signature for functions. We now
introduce some relevant concepts.  Functions can be called with named
parameters, arguments can have default values, and R also support variable
argument lists.  To illustrate all of this in a single example, consider:

\begin{lstlisting}
  f <- function(x, ..., y=if(z==0) 1, z=0) x + y + if (missing(...)) 0 else c(...)
\end{lstlisting}

\noindent
This function four formal parameters, \k x, \k{...}, \k y and \k z. Argument
\k x can be bound positionally or passed by name. The vararg argument,
\k{...}, is always passed positionally. The remaining two arguments must be
passed by name.  Arguments \k y and \k z have default values, in the case of
\k z this is a constant, but \k{y}'s default value is an expression that
depends on the value of \k z. The body of the function will add parameters
\k x and \k y with either the scalar 0 or the result of concatenating the
varargs into a primitive vector. The function \k{missing} tests if an
argument was explicitly passed. The following are some valid invocations
of \k f:

\begin{lstlisting}
 > f(1)    
 [1] 2             % a double vector, y is 0, ... is missing
 > f(1,2) 
 [1] 4             % a double vector, y is 0, ... is 2
 > f(1,2,3)
 [1] 4 5           % a double vector, y is 0, ... is 2, 3
 > f(2,3,x=1)
 [1] 4 5           % a double vector, y is 0, ... is 2, 3
 > f(x=1, y= 1)
 [1] 2             % a double vector, y is 1, ... is missing
 > f(x=1, z= 1)
 numeric(0)        % a double vector of length 0, y is NULL
 > f(1L,2L,y=1L)
 [1] 4             % an integer vector, y is integer 1, ... is integer 2
 > f(1, y=c(1,2))
 [1] 1 2           % a double vetor, y is 1, 2, ... is missing
\end{lstlisting}

\noindent
The above hints at the polymorphism of the language, \k f, may return a
vector of integer or of double, the length of the vector depends on the
length of the varargs and of \k x and \k y.  The language does not really
differentiate between scalar and vectors.
Some more exotic types that can be encountered include vectors of complex
number and list of arbitrary types. Function \k f can be invoked with those
as well.

\begin{lstlisting}
 > c1 <- complex(re=1, im=2)
 > c2 <- complex(re=2, im=1)
 > f(c1)
 [1] 1             % a double vector, x is complex, y is double
 > f(c1,y=c2)
 [1] 2+1i          % a complex vector, x and y are complex
 > l1 <- list(1,2)
 > l2 <- list(c1,c2)
 > f(l1) 
 [1] 1             % a double vector, x is a list of doubles
 > f(l1,y=l2)
 [[1]]             % a list of complex, x is a list of doubles
 [1] 1+2i          %                      y is a list of complex

 [[2]]
 [1] 2+1i
\end{lstlisting}

R has one builtin notion of type that can be queried by the \k{typeof}
function. Figure~\ref{types} lists all of the builtin types that are
provided by the language. They are the possible return values of
\k{typeof}. There is no intrinsic notion of subtyping in R. But, in many
context a \k{logical} will convert to \k{integer}, and an \k{integer} will
convert to \k{double}.  Some odd conversions can occur in corner cases, such
as \k{1<"2"} holds and \k{c(1,2)[1.6]} returns the first element of the
vector, as the double is converted to an integer. R does not distinguish
between scalars and vectors (they are all vectors), so \code{typeof(5) ==}
\code{typeof(c(5)) == typeof(c(5,5))} \code{ == "double"}. Finally all
vectorized data types have a distinguished missing value denoted by
\code{NA}. The default type of \code{NA} is \k{logical}. We can see that
\code{typeof(NA)=="logical"}, but NA inhabits every type:
\code{typeof(c(1,NA)[2])=="double"}.

@@

@@

Scalar data types include the distinguished \k{NULL} value, which is also of
ktype \k{NULL}, instance of classes written using the S4 object system,
closures and environments.  The implementation of R has a number of other
types that are mostly not used by user code, they are listed in
Figure~\ref{types} for reference.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Corpus}\label{sec:corpus} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
